Interactive Visual Analysis of Lumbar Back Pain \subtitle{What the Lumbar Spine Tells About Your Life}

### 2014-10-02 draft.tex
Minor revision of Bernhard

We calculated an correlation matrix to assess correlations between the image variables.

@@

\paragraph{Assessing the Information Gain Using the Principal Component Analysis.}
%%
To determine the information gain per image-derived variable, we calculated a Principal Component Analysis (PCA) and compared the dimension loadings.
%%
The PCA results in a linear projection of the variable space ordering their components by maximum variance.
%%
The first three principal components explain 75\% of the variance in the image-derived variables.
%%
The first principal component explains 47\% of the variance and weights primarily \emph{mean curvature}, \emph{sagittal curvature}, \emph{curvature angle} and \emph{sagittal curvature angle}.
%%
The second component, adding 16\% of the variance, weigths \emph{coronal curvature} and \emph{coronal curvature angle}.
%%
The third component explains 12\% of the variance and weights \emph{torsion} and \emph{transverse curvature}.
%%
This supports our prior conclusion about the low information gain of the transverse planes.
%%
\emph{Torsion} also adds little variance to the information space.
%%
The complete analysis can be found in the experiments section of the supplementary material (\url{blind.dnsalias.com})
%% ------------------------------------------
\paragraph{Heterogenous Correlations.}
%%
We then expanded our focus on correlations of image-derived variables with all other non-image variables.
%%
We applied a heterogenous correlation technique to derive correlations between all variables in the data set.
%%
The method uses the following correlation metrics for the different type combinations:
%%
\begin{itemize}
%%
\item \emph{Pearson product-moment} for two continuous variables, 
\item \emph{Polyserial} correlation for one continuous and one categorical variable, and
\item \emph{Polychoric} correlations for two categorical variables.
\end{itemize}
%%
All correlation values are scaled between \texttt{0 - no correlation} and \texttt{1 - perfect correlation}.
%%
Some variables are too sparse for calculating correlations, for example \emph{treatment of diabetes}, or \emph{medication against high blood pressure}.
%%
These variables have to be omitted in the analysis, since they are not statistically resilient.
%%
We display the resulting \emph{contingency matrix} using a heat map, encoding correlation values with color brightness with white for $0$ and dark blue for $1$, as presented in \cite{Klemm2014VIS}.
%%
We calculated the contingency matrix for all size groups and searched for correlations between image- and non-image derived variables.
%%
The resulting contingency matrices are available as supplementary material and show no strong correlation with any of the variable (see \url{blind.dnsalias.com}).
%%
Only weak correlations could be found for \emph{mean curvature} with \emph{gender} (0.42), \emph{body size} (0.39) and \emph{number of born children} (0.29).
%%
One surprising result was the small correlation of \emph{torsion} with \emph{Parkinson's disease} (0.24).
%%
Other than that, \emph{torsion} correlated with almost no variables (p values between 0 and 0.05).
%%

@@

\section{\uppercase{Evaluation of Decision Trees}}
\label{sec:DecisionTrees}
%%
\noindent As described before, correlation coefficients fail to infer back pain status based on lumbar spine canal \emph{curvature} and \emph{torsion}.
%%
We rely on predictive classification trained to obtain a complex rule set on how combinations of the image-variable explain non-image variables.
%%
Decision trees are a popular classification method in data mining to create predictive models.
%%
%Leafs of a decision tree represent class labels, branches represent decision trees.
Leafs of a decision tree represent class labels, branches represent feature conjunctions leading to the class labels.
%%
Decision trees are easy to understand and to read.
%%
They work with numerical as well as categorical data.
%%
%Epidemiologists can interpret the results without having deep knowledge about the classification algorithm.
%%
Readability is only granted for small trees. %, complex structures with many branches are not desirable.
%%
Too many branches also impose overfitting to the data \cite{DecisionTree}.

\subsection{The C4.5/C5.0 Algorithm}
\label{sec:C5.0}
The C4.5 algorithm builds decision trees based on information entropy on a data set.
%%
Such a calculation requires a numeric or categorical target variable.
%%
%The algorithm then tries to find a decision tree, which divides the samples using the input variables just like the target variables.
%%
%This means, that every node in the tree represents the attribute which splits the data most efficiently into the target subset.
%%
The pseudocode for the algorithm is defined in Algorithm~\ref{alg:c4.5} \cite{C4.5Algorithm}.
\begin{algorithm}[]
 Check for base cases\;
 \For{each attribute $a$}{
  Find the normalized information gain ratio from splitting on $a$\;
 }
 Let $a_{best}$ be the attribute with the highest normalized information gain\;
 Create a decision node that splits on $a_{best}$\;
 Recurse on the sublists obtained by splitting on $a_{best}$, and add those nodes as children of node\;
 \caption{Building a decision tree using the C4.5 Algorithm}
 \label{alg:c4.5}
\end{algorithm}
C5.0 is developed to produce smaller decision trees than C4.5 and improve the execution time.
%%
We use the \texttt{R} implementation of C5.0 \cite{c5.0classification}.
%%
% \begin{enumerate}
% %%
% \item Check for base cases
% %%
% \item For each attribute $a$
% %%
% \begin{enumerate}
% %%
% \item Find the normalized information gain ratio from splitting on $a$
% %%
% \end{enumerate}
% %%
% \item Let $a_{best}$ be the attribute with the highest normalized information gain
% %%
% \item Create a decision node that splits on $a_{best}$
% %%
% \item Recurse on the sublists obtained by splitting on $a_{best}$, and add those nodes as children of node
% %%
% \end{enumerate}
%%
Categorical attributes with more levels are biased with more information gain in a decision tree \cite{deng2011bias}.
%%
%Creating dummy variables bypasses this problem.
Creating a dummy variable by converting each manifestation of a categorical variable into a dichotomous variable bypasses this problem.
%%
%
%The actual use for the resulting tree is the classification of new observations (subjects).
%%
%Yet, we are interested in the decision rules and the classification accuracy.
%%
In the following analysis, we strongly focus on the complexity of decision trees and the classification accuracy.

@@

\paragraph{The Error Term. }
Normally the error for a classification is calculated with $error = \frac{correctlyClassified_{n}} {N}$ where $n$ is the number of subjects.
%%
The metric usually works well for variables with uniform distribution.
%%
It distorts the result for other distribution types.
%%
For example, if a variable indicating a disease is negative for 90\% of the subjects and the classifier simply assigns all subjects to \emph{not ill}, the error metric would yield an error of 10\%, even though it is very bad.
%%
Our error term therefore incorporates the discriminative power of each manifestation and is denoted as follows:
%%
\begin{equation}
error = 1 - \frac{\sum_{m=1}^M \frac{correctlyClassified_{m}}{M}}{number_{dimensions}}
\end{equation}
$M$ represents the set of manifestation of each variable.
%%
The error is scaled to denote perfect classification with 0; 1 is equal to random selection.
%%
We consider a result rated $0.5$ or below good enough to display it in the visualization, a value below $0.25$ represents a good classification.
%%
It allows for comparability of error rates between variables with different manifestation count.

\paragraph{Attribute Mapping}
%%
The scatter plot axes are defined by tree size and the previously described error metric.
%%
This allows us to visualize a multitude of classification results in one plot.
%%
Classification and comparison of variables for subject groups (e.g. male and female subjects) in one plot can be achieved by color coding group affiliation on the data points.

@@

We apply a square root scale for the tree size axis to highlight data points with few decision rules.
%%
Outlier results with very large decision trees would otherwise distort the resulting plot.
%%
\subsubsection{Dummy Variables}
Dummy variables convert a categorical variable with multiple manifestations into several dichotomous variables.
%%
Each dichotomous variable encodes the presence of a manifestation.
%%
For example, a pain indicator variable ranging from \emph{1 - no pain} to \emph{4 - large pain} is subdivided into four dichotomous variables.
%%
One subject can only have one of these variables set to true.
%%
This is useful for our classification, because it allows to determine which manifestation of a variable can be described best using the image data variables.
%%------------
\subsubsection{Interaction With the Visualization}
%%
The described visualization provides a good overview over the classification results.
%%
We still want to be able to display \emph{details-on-demand} \cite{shneiderman1996} and examine a decision tree in detail.
%%
This is realized by clicking on an entry on the visualization, which then interactively displays the corresponding decision tree in detail.
%%
This allows to sequentially analyze the classifications.

@@

\noindent In this section, we show which non-image variables can be described using the 9 image derived variables.

@@

None of the pain indicators can be described reliably using the image-based variables.
%%
The only variable reliably classified in this group is \emph{gender}. % with an error rate of 31.5\% using a decision tree of seven rules.
%
It can be discriminated with an error of $0.31$ using 7 rules and incorporates only \emph{curvature-} and \emph{curvature angle}-related variables.
%%
The distinctness lies in the average difference in \emph{body size} between \emph{male} and \emph{female} subjects.
%%
\emph{Medication for high blood pressure} is classified for 1,058 subjects with an error of $0.47$ solely based on \emph{coronal mean curvature}.
%%
Almost all subjects who are medicated ($796$ of $1,058$) were correctly classified.
%%
The vast majority of non-medicated subjects ($262$ of $1,058$) are false-positive classified, yielding a poor quality of the classifier w.r.t epidemiological research.
%%
The four \emph{body size} groups could be characterized with an error of $0.48$, but the decision tree comprises 71 rules and imposes overfitting.

>>>>

We calculated a correlation matrix to assess correlations between the image variables.

@@

\paragraph{Assessing the Information Gain Using the Principal Component Analysis.}
%%
To determine the information gain per image-derived variable, we calculated a Principal Component Analysis (PCA) and compared the dimension loadings.
%%
The PCA results in a linear projection of the variable space ordering their components by maximum variance.
%%
The first three principal components explain 75\% of the variance in the nine image-derived variables.
%%
The first principal component explains 47\% of the variance and comprises primarily \emph{mean curvature}, \emph{sagittal curvature}, \emph{curvature angle} and \emph{sagittal curvature angle}.
%%
The second component, adding 16\% of the variance, weigths \emph{coronal curvature} and \emph{coronal curvature angle}.
%%
The third component explains 12\% of the variance and comprises \emph{torsion} and \emph{transverse curvature}.
%%
This supports our prior conclusion about the low information gain of the transverse planes.
%%
\emph{Torsion} also adds little variance to the information space.
%%
The complete analysis can be found in the experiments section of the supplementary material (\url{blind.dnsalias.com})
%% ------------------------------------------
\paragraph{Heterogenous Correlations.}
%%
We then expanded our focus on correlations of image-derived variables with all other non-image variables.
%%
%We applied a heterogenous correlation technique to derive correlations between all variables in the data set.
%We applied a heterogenous correlation technique to derive correlations between all variables in the data set.
Different correlation metrics depending on the type of the individual variables are used to derive correlations between all variables in the data set.
%%
The method uses the following correlation metrics for the different type combinations:
%%
\begin{itemize}
%%
\item \emph{Pearson product-moment} for two continuous variables, 
\item \emph{Polyserial} correlation for one continuous and one categorical variable, and
\item \emph{Polychoric} correlations for two categorical variables.
\end{itemize}
%%
All correlation values are scaled between \texttt{0 - no correlation} and \texttt{1 - perfect correlation}.
%%
Some variables are too sparse for calculating correlations, for example \emph{treatment of diabetes}, or \emph{medication against high blood pressure}.
%%
These variables have to be omitted in the analysis, since they are not statistically resilient.
%%
We display the resulting \emph{contingency matrix} using a heat map, encoding correlation values with color brightness with white for $0$ and dark blue for $1$, as presented in \cite{Klemm2014VIS}.
%%
We calculated the contingency matrix for all size groups and searched for correlations between image- and non-image derived variables.
%%
The resulting contingency matrices are available as supplementary material and show no strong correlation with any of the variable (see \url{blind.dnsalias.com}).
%%
Only weak correlations could be found for \emph{mean curvature} with \emph{gender} (0.42), \emph{body size} (0.39) and \emph{number of born children} (0.29).
%%
One surprising result was the small correlation of \emph{torsion} with \emph{Parkinson's disease} (0.24).
%%
Other than that, \emph{torsion} correlated with almost no variables (values between 0 and 0.05).
%%

@@

\section{\uppercase{Evaluation of Decision Trees}}
\label{sec:DecisionTrees}
%%
\noindent As described before, correlation coefficients fail to infer back pain status based on lumbar spine canal \emph{curvature} and \emph{torsion}.
%%
We rely on predictive classification trained to obtain a complex rule set on how combinations of the image-variable explain non-image variables.
%%
Decision trees are a established classification method in data mining to create predictive models.
%%
%Leafs of a decision tree represent class labels, branches represent decision trees.
Leafs of a decision tree represent class labels, branches represent feature conjunctions leading to the class labels.
%%
Decision trees are easy to understand and to read.
%%
They work with numerical as well as categorical data.
%%
%Epidemiologists can interpret the results without having deep knowledge about the classification algorithm.
%%
Readability is only granted for small trees. %, complex structures with many branches are not desirable.
%%
Large, complex trees are difficult to trace and to understand.
%%
Too many branches also impose overfitting to the data \cite{DecisionTree}.

\subsection{The C4.5/C5.0 Algorithm}
\label{sec:C5.0}
The C4.5 algorithm builds decision trees based on information entropy on a data set.
%%
Such a calculation requires a numeric or categorical target variable.
%%
%The algorithm then tries to find a decision tree, which divides the samples using the input variables just like the target variables.
%%
%This means, that every node in the tree represents the attribute which splits the data most efficiently into the target subset.
%%
The pseudocode for the algorithm is defined in Algorithm~\ref{alg:c4.5} \cite{C4.5Algorithm}.
\begin{algorithm}[]
 Check for base cases\;
 \For{each attribute $a$}{
  Find the normalized information gain ratio from splitting on $a$\;
 }
 Let $a_{best}$ be the attribute with the highest normalized information gain\;
 Create a decision node that splits on $a_{best}$\;
 Iterate on the sublists obtained by splitting on $a_{best}$, and add those nodes as children of node\;
 \caption{Building a decision tree using the C4.5 Algorithm}
 \label{alg:c4.5}
\end{algorithm}
C5.0 is developed to produce smaller decision trees than C4.5 and improve the execution time.
%%
We use the \texttt{R} implementation of C5.0 \cite{c5.0classification}.
%%
% \begin{enumerate}
% %%
% \item Check for base cases
% %%
% \item For each attribute $a$
% %%
% \begin{enumerate}
% %%
% \item Find the normalized information gain ratio from splitting on $a$
% %%
% \end{enumerate}
% %%
% \item Let $a_{best}$ be the attribute with the highest normalized information gain
% %%
% \item Create a decision node that splits on $a_{best}$
% %%
% \item Recurse on the sublists obtained by splitting on $a_{best}$, and add those nodes as children of node
% %%
% \end{enumerate}
%%
Categorical attributes with more levels are biased with more information gain in a decision tree \cite{deng2011bias}.
%%
%Creating dummy variables bypasses this problem.
Creating a dummy variable by converting each manifestation of a categorical variable into a dichotomous variable bypasses this problem.
%%
%
%The actual use for the resulting tree is the classification of new observations (subjects).
%%
%Yet, we are interested in the decision rules and the classification accuracy.
%%
In the following analysis, we strongly focus on the complexity of decision trees and the classification accuracy.

@@

\paragraph{The Error Term. }
Normally the error for a classification is calculated with $error = \frac{correctlyClassified_{n}} {N}$ where $n$ is the number of subjects.
%%
The metric usually works well for variables with uniform distribution.
%%
It distorts the result for other distribution types.
%%
For example, if a variable indicating a disease is negative for 90\% of the subjects and the classifier simply assigns all subjects to \emph{not ill}, the error metric would yield an error of 10\%, even though it is very bad.
%%
Based on this we use a summary error, which incorporates the discriminative power of each manifestation and is denoted as follows:
%Our error term therefore incorporates the discriminative power of each manifestation and is denoted as follows:
%%
\begin{equation}
error = 1 - \frac{\sum_{m=1}^M \frac{correctlyClassified_{m}}{M}}{number_{dimensions}}
\end{equation}
$M$ represents the set of manifestation of each variable.
%%
The error is scaled to denote perfect classification with 0; 1 is equal to random selection.
%%
We consider a result rated $0.5$ or below good enough to display it in the visualization, a value below $0.25$ represents a good classification.
%%
It allows for comparability of error rates between variables with different manifestation count.

\paragraph{Attribute Mapping.}
%%
The scatter plot axes are defined by tree size and the previously described error metric.
%%
This allows us to visualize a multitude of classification results in one plot.
%%
Classification and comparison of variables for subject groups (e.g. male and female subjects) in one plot can be achieved by color coding group affiliation on the data points.

@@

We apply a square root scale for the tree size axis to highlight decision trees with few decision rules.
%%
Outlier results with very large decision trees would otherwise distort the resulting plot.
%%
\subsubsection{Dummy Variables}
Dummy variables convert a categorical variable with multiple manifestations into several dichotomous variables.
%%
Each dichotomous variable encodes the presence of a manifestation.
%%
For example, a pain indicator variable ranging from \emph{1 - no pain} to \emph{4 - large pain} is subdivided into four dichotomous variables.
%%
One subject can only have one of these variables set to true.
%%
This is useful for our classification, because it allows to determine which manifestation of a variable can be described best using the image data variables.
%%------------
\subsubsection{Interaction With the Visualization}
%%
The described visualization provides a good overview over the classification results.
%%
We still want to be able to display \emph{details-on-demand} \cite{shneiderman1996} and examine a decision tree in detail.
%%
This is realized by clicking on an entry in the visualization, which then interactively displays the corresponding decision tree in detail.
%%
This allows to sequentially analyze the classifications.

@@

\noindent In this section, we show which non-image variables can be explained using the 9 image derived variables.

@@

None of the pain indicators can be described reliably using the image-based variables.
%%
The only variable reliably classified in this group is \emph{gender}, which can be predicted with an error of $0.31$ using 7 rules and incorporates only \emph{curvature-} and \emph{curvature angle}-related variables.
%
%It can be predicted with an error of $0.31$ using 7 rules and incorporates only \emph{curvature-} and \emph{curvature angle}-related variables.
%%
The distinctness lies in the average difference in \emph{body size} between \emph{male} and \emph{female} subjects.
%%
\emph{Medication for high blood pressure} is classified for 1,058 subjects with an error of $0.47$ solely based on \emph{coronal mean curvature}.
%%
Almost all subjects who are medicated ($796$ of $1,058$) were correctly classified.
%%
The vast majority of non-medicated subjects ($262$ of $1,058$) are false-positive classified, yielding a poor quality of the classifier w.r.t epidemiological research.
%%
The four \emph{body size} groups could be characterized with an error of $0.48$, but the decision tree comprises 71 rules and imposes overfitting.
